Environmental and geographic variables are effective surrogates for genetic variation in conservation planning

### 2017-01-22 code/rmarkdown/article.Rmd
assorted improvements to article main text

---
title: "Environmental and geographic variables are effective surrogates for genetic variation in conservation planning"
author:
  - name: Jeffrey O. Hanson
    affiliation: a,1
  - name: Jonathan R. Rhodes
    affiliation: b
  - name: Cynthia Riginos
    affiliation: a
  - name: Hugh P. Possingham
    affiliation: a
  - name: Richard A. Fuller
    affiliation: a
address:
  - code: a
    address: School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  - code: b
    address: School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
corresponding_author:
  - code: 1
  - text: "To whom correspondence should be addressed. E-mail: jeffrey.hanson@uqconnect.edu.au"
author_contributions: |
  All authors developed the study and drafted the manuscript. JOH performed all analyses.
lead_author_surname: "Hanson"
abstract: |
  Protected areas buffer species from anthropogenic threats and provide places for processes to continue. However, evolutionary processes are not often explicitly considered when deciding where to site protected areas because genetic data require considerable resources to obtain and analyse. Here we show that environmental and geographic distance variables are effective surrogates for adaptive and neutral intra-specific variation for conservation planning. We obtained occurrence and genetic data from the IntraBioDiv project for 27 species collected over the European Alps using a gridded sampling scheme. For each species, we identified loci that were potentially under selection using outlier loci methods, and mapped their main gradients of adaptive and neutral genetic variation across the grid cells. We then used the cells as planning units to generate prioritisations. Our first analysis verified that the spatial patterns of environmental and geographic variation correlated with those adaptive and neutral genetic variation (respectively). Our second analysis showed that these surrogates can predict the proportion of genetic variation secured in randomly generated solutions. Our third analysis found that solutions generated using additional surrogate-based targets secured more neutral genetic variation. We discovered that solutions generated using traditional amount-based targets tend to secure most species' adaptive genetic variation. Overall, our work provides a rigorous examination of two potential surrogates for genetic variation.
significance: |
  Protected areas are used to preserve biodiversity patterns and processes. Since the resources available for conservation are limited, protected areas need to be sited in cost-effective locations. However, despite their importance, evolutionary processes are not often used to guide reserve selection. This is primarily because genetic data require extensive resources to obtain and process. Here we show that freely available environmental and geographic variables can be used as surrogates for genetic data in conservation planning.
acknowledgements: |
  JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Biodiversity and Conservation Science (CBCS).
keywords:
  - conservation
  - biodiversity
  - AFLP
  - genetic diversity
pnas_type: pnasresearcharticle
bibliography: references.bib
csl: reference-style.csl
output:
  rticles::pnas_article:
    includes:
      in_header: [preamble.tex]
---

@@

Recently, the use of surrogate data has been proposed to generate prioritisations that sufficiently intra-specific genetic variation--without needing to directly utilise genetic data to achieve this. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been proposed to guide reserve selection as surrogates for adaptive genetic variation (eg. [@r202; @r2]). By generating prioritisations that preserve a more representative sample of environmental conditions across species' ranges, these prioritisations could also capture more adaptive variation. However, the effectiveness of environmental surrogates in representing adaptive genetic variation remains unverified. Neutral genetic variation arises from a reduction in gene flow between individuals. Over the last few years, the field of phylogeography has predominantly focused on describing neutral genetic diversity, and so the landscape factors that affect neutral variation are relatively well understood (reviewed in [@r501]). Potential surrogates for neutral variation have been based on variables that predict the levels of connectivity between planning units. For instance, isolation by distance (IBD; [@r259]) predicts that populations further apart will experience less gene flow, and in turn, share less neutral loci. Based on this idea, geographic surrogates based on distance have been found to improve conservation prioritisations in an island network ([@r200]). However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors (eg. habitat; [@r497])--the typical situations in most conservation planning exercises.

Here we investigate the effectiveness of environmental and geographic surrogates for capturing adaptive and neutral genetic variation respectively in the context of conservation planning. We used distribution and genomic data for 27 alpine plants collected by the international IntraBioDiv project in the European Alps ([@r451]). We show that for most species the spatial patterns of adaptive and neutral genetic variation correlated with environmental and geographic variables. We also show that for most species the level of association is strong enough to be operational for conservation planning. Finally, we demonstrate the using these surrogates results in prioritisations that secure more genetic variation than expected by chance.

@@

<!-- summary of methods -->
We detected putatively adaptive genetic variation in `r as.character(as.english(length(env.surrogacy.LST)))` of the 27 plant species. We used two outlier detection methods to detect loci under selection, and they returned reasonably consistent results (mean `r round(mean(spp.outlier.methods.congruence, na.rm=TRUE),2)` % loci per species assigned the same classification $\pm$ `r round(sd(spp.outlier.methods.congruence, na.rm=TRUE),2)` SD; Table S1). Of the species that were associated with adaptive genetic variation, only a small proportion of loci were classified as being adaptive (mean `r round(mean(spp.outlier.loci.props, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.loci.props, na.rm=TRUE),2)` SD). After identifying the loci showing strong signals of selection, we used non-metric multi-dimensional scaling (NMDS) analyses to identify the main gradients of adaptive (if detected) and neutral variation for each species. Generally, only a small number of continuous dimensions was needed to sufficiently describe their patterns of adaptive (all $K = `r unique(na.omit(adaptive.k))`$; Table S2) and neutral genetic variation ($K = `r min(neutral.k)`$--$`r max(neutral.k)`$; Table S1). The resulting ordinations were used to construct an adaptive (if detected) and neutral genetic space for each species. The spatial distribution of these genetic spaces (Figures S29--S54) generally showed substantial spatial auto-correlation (eg. _Cerastium uniflorum_, Figure S34; _Dryas octopetala_, Figures S36), suggesting that the environmental and geographic variables may prove to be effective surrogates.

<!-- correlation analysis -->
We verified that the spatial patterns in environmental and geographic variation correlated with the broad-scale spatial patterns in adaptive and neutral variation for most species (respectively). For each species,we constructed dissimilarity matrices based on the planning units they were found in. These matrices expressed differences between the planning units in terms of their (i) environmental characteristics, (ii) geographic position, and also the (iii) adaptive (if detected), and (iv) neutral genetic characteristics of individuals found inside them. By correlating these matrices, the spatial patterns of environmental variation were found to be significantly correlated with that of adaptive genetic variation for `r as.character(as.english(sum(correlation.model.summary.DF$Environmental.corr.P<0.05,na.rm=TRUE)))` of the `r as.character(as.english(sum(!is.na(correlation.model.summary.DF$Environmental.corr.P))))` species associated with adaptive variation (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` SD for significant models; Table S3). Similarly, geographic distance between planning units was also correlated significantly with the spatial pattern of neutral genetic variation among planning units for `r sum(correlation.model.summary.DF$Geographic.corr.P<0.05,na.rm=TRUE)` of the `r sum(!is.na(correlation.model.summary.DF$Geographic.corr.P))` species (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` SD for significant models; Table S3). Thus, for most species, planning units that contained different environmental conditions contained individuals with different adaptive genetic characteristics, and planning units that were further apart tended to contain individuals with different neutral genetic characteristics. After verifying these patterns were correlated, our next step was to determine whether they were correlated enough to be effective for conservation planning.

<!-- surrogacy analysis -->
For each species, we generated a suite of `r format(raptr.params.LST[[MODE]]$surrogacy.analysis$random.replicates, big.mark=',', scientific=FALSE)` random prioritisations, and calculated the proportion of environmental variation and adaptive (if present) genetic variation they sampled. We then repeated this process, and calculated the proportion of geographic variation and neutral genetic variation they sampled. The environmental and geographic variables were moderately effective predictors for the genetic variation represented by the randomly generated prioritisations (Figure 1). The relationship between the proportion of genetic and surrogate variation secured in the prioritisations varied among the species (species $\times$ proportion of environmental variation interaction term: ${\chi^2}_{`r env.surrogacy.AOV$df[[4]]`} = {`r round(env.surrogacy.AOV$Resid..Dev[[4]],2)`}; P `r pvalString(env.surrogacy.AOV$Pr..Chi.[[4]])`$; species $\times$ proportion of geographic variation interaction term: ${\chi^2}_{`r round(geo.surrogacy.AOV$df[[4]],2)`} = `r round(geo.surrogacy.AOV$Resid..Dev[[4]],2)`$; $P$ `r pvalString(geo.surrogacy.AOV$Pr..Chi.[[4]])`). _Post-hoc_ analyses showed that these relationships were positive for all species (environmental: minimum \textit{z} = `r round(max(env.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; geographic: minimum \textit{z} = `r round(min(geo.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S4). After establishing that the the proportion of surrogate variation secured in a prioritisation also predicted the proportion of genetic variation it secured, our final step was to determine if prioritisations were more effective when generated using targets based on the surrogates.

@@

The average proportion of genetic variation secured by prioritisations varied under different scenarios (${\chi^2}_{`r scenario.ANOVA.2.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.2.DF$statistic[[2]],2)`$; _P_ `r pvalString(scenario.ANOVA.2.DF$p.value[[2]])`). Both the multi-species (mean `r round(mean(filter(scenario.sub.DF,Context=='Multi-species\n(equal costs)')$genetic.held*100),2)` % genetic variation secured $\pm$ `r round(sd(filter(scenario.sub.DF,Context=='Multi-species\n(equal costs)')$genetic.held*100),2)`) and multi-species with cost (mean `r round(mean(filter(scenario.sub.DF,Context=='Multi-species\n(opportunity costs)')$genetic.held*100),2)` % genetic variation secured $\pm$ `r round(sd(filter(scenario.sub.DF,Context=='Multi-species\n(opportunity costs)')$genetic.held*100),2)`) were both found to secure more genetic variation than the single species scenario (mean `r round(mean(filter(scenario.sub.DF,Context=='Single-species\n(equal costs)')$genetic.held*100),2)` % genetic variation secured $\pm$ `r round(sd(filter(scenario.sub.DF,Context=='Single-species\n(equal costs)')$genetic.held*100),2)`; $Z_1 = `r round(scenario.GLHT.1.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.GLHT.1.DF$p.value[[2]])`$; $Z_1 = `r round(scenario.GLHT.1.DF$statistic[[1]],2)`$; $P = `r pvalString(scenario.GLHT.1.DF$p.value[[1]])`$ respectively). These results suggest that on average the multi-species prioritisations can secure more genetic variation than single-species prioritisations--regardless of targets used to generate them.

@@

<!--  introduction -->
We aimed to investigate the effectiveness of broad-scale environmental and geographic variables as surrogates for adaptive and neutral genetic variation in conservation planning. Our study is the first to address this aim using a broad range of species and calibrated against field measurements of genetic variation. The spatial patterns of environmental and geographic surrogates were strongly correlated with the patterns of adaptive and neutral genetic variation for most species, and environmental and geographic targets resulted in prioritisations that secured a greater proportion of adaptive and neutral genetic variation than traditional conservation planning for most species. These environmental and geographic surrogates are based on freely available data, and could be calculated for any study region across the world.

<!-- surrogate/genetic held correlations -->
Our results demonstrate that environmental and geographic surrogates are strong enough to be operational (Figure 2). For most species, the environmental and geographic surrogates variables were effective surrogates for adaptive (mean `r round(mean(env.surrogate.spp.DF$r2CU),2)` $R^2$ $\pm$ `r round(sd(env.surrogate.spp.DF$r2CU),2)` SD; Figures 2c and S56) and neutral genetic variation (mean `r round(mean(geo.surrogate.spp.DF$r2CU),2)` $R^2$ $\pm$ `r round(sd(geo.surrogate.spp.DF$r2CU),2)` SD; Figures 2d and Figure S57). Despite this, for a few species, geographic distance was a poor surrogate for neutral genetic variation (eg. _Gypsophila repens_: $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Gypsophila_repens')$Geographic.r2CU,2)`; _Luzula alpinopilosa_: $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Luzula_alpinopilosa')$Geographic.r2CU,2)`; _Gentiana nivalis_ $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Gentiana_nivalis')$Geographic.r2CU,2)`). Indeed, geographic distances will be a poor surrogate for species that utilise dispersal modes that result in non-linear relationships between connectivity and geographic distance [eg. wind dispersing plants can maintain high connectivity between distant areas; @r451], and in landscapes where connectivity is most affected by landscape features ([@r497]). Overall, however, such species were the exception in our analysis, and both simple surrogates performed moderately well for a diverse set of species.

<!-- suitable targets for surrogates -->
Our results suggest that environmental and geographic surrogates can predict how much genetic variation prioritisations secure, but what targets should be used? First, our results suggest that surrogate-based targets should be used in conjunction with amount-based targets. For all species, it was possible to generate solutions that secured a high proportion of the surrogate variable (> 90 %) and only a small proportion of adaptive or neutral genetic variation (< 20 %; Figures S56--S57), and so use of amount-based targets may ameliorate this. Second, geographic surrogates may need much higher targets than environmental surrogates to capture evolutionary processes. In our analysis, geographic variation accumulated much slower for neutral genetic representation than environmental variation did with adaptive representation (Figures 2a--2b and S56--S57). To secure a representative sample of the species' neutral genetic variation ($\geq$ `r raptr.params.LST[[MODE]]$scenario.analysis$genetic.target*100` %), prioritisations needed to sample `r round(mean(surrogate.repr.prop.DF$geographic.held*100,na.rm=TRUE),2)` % $\pm$ `r round(sd(surrogate.repr.prop.DF$geographic.held*100,na.rm=TRUE),2)` SD of the variation in geographic position among the planning units occupied by the species. On the other hand, adaptive variation accumulated much faster. To secure at least `r raptr.params.LST[[MODE]]$scenario.analysis$genetic.target*100` % of each species' adaptive genetic variation, prioritisations only needed to sample on average `r round(mean(surrogate.repr.prop.DF$environmental.held*100,na.rm=TRUE),2)` % $\pm$ `r round(sd(surrogate.repr.prop.DF$environmental.held*100,na.rm=TRUE),2)` SD of the environmental conditions found across the planning units occupied by the species. Note that these proportions are based on averages across species. In many cases, higher targets may be needed to ensure that prioritisations secure a representative sample of the genetic variation for all species in a planning exercise.

<!-- adaptive genetic variation secured in prioritisations-->
Our results suggest that conservation planners can secure a reasonably representative sample of the intra-specific adaptive genetic variation using only amount-based targets. For most species, the spatial patterns of adaptive genetic variation clustered into a few distinct groups (eg. _Carex semipervens_ and _Cerastium uniflorum_; Figures S33 and S34 respectively). As a consequence, a prioritisation would only need to select a planning unit in each cluster to secure a representative sample. This mechanism explains why adaptive variation accumulated much quicker than neutral genetic variation in the randomly generated prioritisations. Additionally, this mechanism also explains why the single-species prioritisations generated using only amount targets were often able to achieve high representation--they were essentially a random selection and so frequently contained planning units in different parts of the species' ranges (Figure 4a). Overall--and contrary to our original hypotheses--it seems that conventional reserve selection methods may secure a representative sample of species' adaptive genetic variation.

<!--  limitations/ weaknesses -->
There are several limitations associated with our analysis. First, the size of the planning units we used (23 $\times$ 25 km) is much larger than typically used in regional conservation planning scenarios (1--10 km$^2$). We used this resolution because the genetic data were collected at this scale. Whilst we could have interpolated the genetic data to a finer resolution and used smaller planning units, this would have biased our analysis by artificially introducing additional spatial auto-correlation. Second, we used geographic distances as surrogates for neutral genetic variation, but connectivity in this landscape may be better described using distances that incorporate information on topography. However, such distances require specific-specific scaling to adequate describe their connectivity (eg. [@r497]), and so cannot easily be utilised in large-scale planning exercises. Third, we used amplified fragment length polymorphism data (AFLP; [@r453]) to characterise the main gradients of genetic variation. Whilst next-generation sequencing provides higher resolution genetic information than AFLP data ([@r480]), no multi-species genomic data set with similar properties currently exists, and our methodology would still have utilised only the main gradients of the genetic variation to generate prioritisations in a feasible amount of time.

<!--  conclusion -->
Broad-scale environmental and geographic variables are generally effective surrogates for representing adaptive and neutral genetic variation for most species in our study system. Moreover, data to calculate the surrogate variables are freely available for any location in the world. Careful use of such surrogates in conservation planning could vastly improve the chances of long-term biodiversity persistence for relatively additional cost.

>>>>

---
title: "Environmental and geographic variables are effective surrogates for genetic variation in conservation planning"
author:
  - name: Jeffrey O. Hanson
    affiliation: a,1
  - name: Jonathan R. Rhodes
    affiliation: b
  - name: Cynthia Riginos
    affiliation: a
  - name: Hugh P. Possingham
    affiliation: a
  - name: Richard A. Fuller
    affiliation: a
address:
  - code: a
    address: School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  - code: b
    address: School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
corresponding_author:
  - code: 1
  - text: "To whom correspondence should be addressed. E-mail: jeffrey.hanson@uqconnect.edu.au"
author_contributions: |
  All authors developed the study and drafted the manuscript. JOH performed all analyses.
lead_author_surname: "Hanson"
abstract: |
  Protected areas buffer species from anthropogenic threats and provide places for biodiversity processes to continue. However, evolutionary processes are not often explicitly considered when deciding where to site protected areas because genetic data require considerable resources to obtain and analyse. Here we show that environmental and geographic distance variables are effective surrogates for adaptive and neutral intra-specific variation for conservation planning. We obtained occurrence and genetic data from the IntraBioDiv project for 27 alpine plant species collected over the European Alps using a gridded sampling scheme. For each species, we identified loci that were potentially under selection using outlier loci methods, and mapped their main gradients of adaptive and neutral genetic variation across the grid cells. We then used the cells as planning units to generate prioritisations. Our first analysis verified that the spatial patterns of environmental and geographic variation correlated with those adaptive and neutral genetic variation (respectively). Our second analysis showed that these surrogates can predict the proportion of genetic variation secured in randomly generated solutions. Our third analysis found that solutions generated using additional surrogate-based targets secured more neutral genetic variation. We discovered that solutions generated using traditional amount-based targets tend to secure most species' adaptive genetic variation. Overall, our work provides a rigorous examination of two potential surrogates for genetic variation.
significance: |
  Protected areas are used to preserve biodiversity patterns and processes. Since the resources available for conservation are limited, protected areas need to be sited in cost-effective locations. However, despite their importance, evolutionary processes are not often used to guide reserve selection. This is primarily because genetic data require extensive resources to obtain and process. Here we show that freely available environmental and geographic variables can be used as surrogates for genetic data in conservation planning.
acknowledgements: |
  JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Biodiversity and Conservation Science (CBCS).
keywords:
  - conservation
  - biodiversity
  - AFLP
  - genetic diversity
pnas_type: pnasresearcharticle
bibliography: references.bib
csl: reference-style.csl
output:
  rticles::pnas_article:
    includes:
      in_header: [preamble.tex]
---

@@

Recently, the use of surrogate data has been proposed to generate prioritisations that sufficiently capture intra-specific genetic variation--without needing to directly utilise genetic data to achieve this. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been proposed to guide reserve selection as surrogates for adaptive genetic variation (eg. [@r202; @r2]). By generating prioritisations that preserve a more representative sample of environmental conditions across species' ranges, these prioritisations could also capture more adaptive variation. However, the effectiveness of environmental surrogates in representing adaptive genetic variation remains unverified. Neutral genetic variation arises from a reduction in gene flow between individuals. Over the last few years, the field of phylogeography has predominantly focused on describing neutral genetic diversity, and so the landscape factors that affect neutral variation are relatively well understood (reviewed in [@r501]). Potential surrogates for neutral variation have been based on variables that predict the levels of connectivity between planning units. For instance, isolation by distance (IBD; [@r259]) predicts that populations located further apart will experience less gene flow, and in turn, share less neutral loci. Based on this idea, geographic surrogates based on distance have been found to improve conservation prioritisations in an island network ([@r200]). However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors (eg. habitat; [@r497])--the typical situations in most conservation planning exercises.

Here we investigate the effectiveness of environmental and geographic surrogates for capturing adaptive and neutral genetic variation respectively in the context of conservation planning. We used distribution and genomic data for 27 alpine plant species in the European Alps that was obtained by the IntraBioDiv project ([@r451]). The data was collected following a gridded sampling scheme which we adopted as planning units for this study. We show that for most species the spatial patterns of adaptive and neutral genetic variation correlated with environmental and geographic variables. We also show that for most species the level of association is strong enough to be operational for conservation planning. Finally, we demonstrate that using these surrogates results in prioritisations that secure more genetic variation than expected by chance.

@@

<!-- summary of methods -->
We detected putatively adaptive genetic variation in `r as.character(as.english(length(env.surrogacy.LST)))` of the 27 plant species. We used two outlier detection methods to identify loci under selection, and they returned reasonably consistent results (mean `r round(mean(spp.outlier.methods.congruence, na.rm=TRUE),2)` % loci per species assigned the same classification $\pm$ `r round(sd(spp.outlier.methods.congruence, na.rm=TRUE),2)` SD; Table S1). Of the species that were associated with adaptive genetic variation, only a small proportion of loci were classified as being adaptive (mean `r round(mean(spp.outlier.loci.props, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.loci.props, na.rm=TRUE),2)` SD). After identifying the loci showing strong signals of selection, we used non-metric multi-dimensional scaling (NMDS) analyses to identify the main gradients of adaptive (if detected) and neutral variation for each species. Generally, only a small number of continuous dimensions was needed to sufficiently describe their patterns of adaptive ($K = `r unique(na.omit(adaptive.k))`$ for all species; Table S2) and neutral genetic variation ($`r min(neutral.k)` \leq K \leq `r max(neutral.k)`$ for all species; Table S1). The resulting ordinations were used to construct an adaptive (if detected) and neutral genetic space for each species. The spatial distribution of these genetic spaces (Figures S29--S54) generally showed substantial spatial auto-correlation (eg. _Cerastium uniflorum_, Figure S34; _Dryas octopetala_, Figure S36), suggesting that the environmental and geographic variables may prove to be effective surrogates.

<!-- correlation analysis -->
We verified that the spatial patterns in environmental variation correlated with the broad-scale patterns in adaptive genetic variation. We further verified that there was a correlation between neutral genetic variation and variation in geographic position. For each species, we constructed dissimilarity matrices expressing differences between the planning units within which they were detected according to terms their (i) environmental characteristics, (ii) geographic position, and also the (iii) adaptive (if adaptive loci detected), and (iv) neutral genetic characteristics of individuals found inside them. By correlating these matrices, the spatial patterns of environmental variation were found to be significantly correlated with that of adaptive genetic variation for `r as.character(as.english(sum(correlation.model.summary.DF$Environmental.corr.P<0.05,na.rm=TRUE)))` of the `r as.character(as.english(sum(!is.na(correlation.model.summary.DF$Environmental.corr.P))))` species associated with adaptive variation (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` SD for significant models; Table S3). Similarly, geographic distance between planning units was also correlated significantly with the spatial pattern of neutral genetic variation among planning units for `r sum(correlation.model.summary.DF$Geographic.corr.P<0.05,na.rm=TRUE)` of the `r sum(!is.na(correlation.model.summary.DF$Geographic.corr.P))` species (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` SD for significant models; Table S3). Thus, for most species, planning units that contained different environmental conditions contained individuals with different adaptive genetic characteristics, and planning units that were further apart tended to contain individuals with different neutral genetic characteristics. After verifying these patterns were correlated, our next step was to determine whether they were correlated enough to be effective for conservation planning.

<!-- surrogacy analysis -->
For each species, we generated a suite of `r format(raptr.params.LST[[MODE]]$surrogacy.analysis$random.replicates, big.mark=',', scientific=FALSE)` random prioritisations, and calculated the proportion of environmental variation and adaptive (if present) genetic variation each prioritisation captured. We then repeated this process, and calculated the proportion of geographic variation and neutral genetic variation they sampled. The environmental and geographic variables were moderately effective predictors for the genetic variation represented by the randomly generated prioritisations (Figure 1). The relationship between the proportion of genetic and surrogate variation secured in the prioritisations varied among the species (species $\times$ proportion of environmental variation interaction term: ${\chi^2}_{`r env.surrogacy.AOV$df[[4]]`} = {`r round(env.surrogacy.AOV$Resid..Dev[[4]],2)`}; P `r pvalString(env.surrogacy.AOV$Pr..Chi.[[4]])`$; species $\times$ proportion of geographic variation interaction term: ${\chi^2}_{`r round(geo.surrogacy.AOV$df[[4]],2)`} = `r round(geo.surrogacy.AOV$Resid..Dev[[4]],2)`$; $P$ `r pvalString(geo.surrogacy.AOV$Pr..Chi.[[4]])`). _Post-hoc_ analyses showed that these relationships were positive for all species (environmental: minimum \textit{z} = `r round(max(env.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; geographic: minimum \textit{z} = `r round(min(geo.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S4). After establishing that the the proportion of surrogate variation secured in a prioritisation also predicted the proportion of genetic variation it secured, our final step was to determine if prioritisations were more effective when generated using targets based on the surrogates.

@@

The average proportion of genetic variation secured by prioritisations varied under different scenarios (${\chi^2}_{`r scenario.ANOVA.2.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.2.DF$statistic[[2]],2)`$; _P_ `r pvalString(scenario.ANOVA.2.DF$p.value[[2]])`). Both the multi-species (mean `r round(mean(filter(scenario.sub.DF,Context=='Multi-species\n(equal costs)')$genetic.held*100),2)` % genetic variation secured $\pm$ `r round(sd(filter(scenario.sub.DF,Context=='Multi-species\n(equal costs)')$genetic.held*100),2)`) and multi-species with cost (mean `r round(mean(filter(scenario.sub.DF,Context=='Multi-species\n(opportunity costs)')$genetic.held*100),2)` % genetic variation secured $\pm$ `r round(sd(filter(scenario.sub.DF,Context=='Multi-species\n(opportunity costs)')$genetic.held*100),2)`) were found to secure more genetic variation than the single species scenario (mean `r round(mean(filter(scenario.sub.DF,Context=='Single-species\n(equal costs)')$genetic.held*100),2)` % genetic variation secured $\pm$ `r round(sd(filter(scenario.sub.DF,Context=='Single-species\n(equal costs)')$genetic.held*100),2)`; $Z_1 = `r round(scenario.GLHT.1.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.GLHT.1.DF$p.value[[2]])`$; $Z_1 = `r round(scenario.GLHT.1.DF$statistic[[1]],2)`$; $P = `r pvalString(scenario.GLHT.1.DF$p.value[[1]])`$ respectively). These results suggest that on average the multi-species prioritisations can secure more genetic variation than single-species prioritisations--regardless of targets used to generate them.

@@

<!--  introduction -->
We aimed to investigate the effectiveness of broad-scale environmental and geographic variables as surrogates for adaptive and neutral genetic variation in conservation planning. Our study is the first to address this aim using field measurements of genetic variation for a broad range of species to assess the performance of prioritisations. The spatial patterns of environmental and geographic surrogates were strongly correlated with the patterns of adaptive and neutral genetic variation for most species. Moreover, for most species, prioritisations generated using environmental and geographic targets secured a greater proportion of adaptive and neutral genetic variation than traditional conservation planning. These environmental and geographic surrogates were based on freely available data, and could be applied to any study region across the world.

<!-- surrogate/genetic held correlations -->
Our results demonstrate that environmental and geographic surrogates are strong enough to be operational for conservation planning (Figure 2). For most species, the environmental and geographic surrogates variables were effective surrogates for adaptive (mean `r round(mean(env.surrogate.spp.DF$r2CU),2)` $R^2$ $\pm$ `r round(sd(env.surrogate.spp.DF$r2CU),2)` SD; Figures 2c and S56) and neutral genetic variation (mean `r round(mean(geo.surrogate.spp.DF$r2CU),2)` $R^2$ $\pm$ `r round(sd(geo.surrogate.spp.DF$r2CU),2)` SD; Figures 2d and Figure S57). Despite this, for a few species, geographic distance was a poor surrogate for neutral genetic variation (eg. _Gypsophila repens_: $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Gypsophila_repens')$Geographic.r2CU,2)`; _Luzula alpinopilosa_: $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Luzula_alpinopilosa')$Geographic.r2CU,2)`; _Gentiana nivalis_ $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Gentiana_nivalis')$Geographic.r2CU,2)`). Indeed, geographic distances are expected to be a poor surrogate for connectivity when species deviate from isolation by distance (eg. wind dispersing plants can maintain high connectivity between distant areas; [@r451]), and in areas where connectivity is most affected by landscape features ([@r497]). Overall, such species were the exception in our analysis, and both surrogates performed moderately well for most species.

<!-- suitable targets for surrogates -->
Our results suggest that environmental and geographic surrogates can be used to capture genetic variation in prioritisations, the next question is: what percentage of surrogate variation should planners target to preserve genetic variation? Firstly, our results suggest that surrogate-based targets should be used with amount-based targets. For all species, it was possible to generate solutions that secured a high proportion of the surrogate variable (> 90 %) and only a small proportion of genetic variation (< 20 %; Figures S56--S57). By using both amount- and surrogate-based targets, planners may avoid such solutions. Secondly, geographic surrogates may need higher targets than environmental surrogates to capture evolutionary processes. In our analysis, geographic variation accumulated at a slower rate for neutral genetic representation than environmental variation did with adaptive representation (Figures 2a--2b and S56--S57). To secure a representative sample of each species' neutral genetic variation ($\geq$ `r raptr.params.LST[[MODE]]$scenario.analysis$genetic.target*100` %), prioritisations needed to sample at least `r round(mean(surrogate.repr.prop.DF$geographic.held*100,na.rm=TRUE),2)` % $\pm$ `r round(sd(surrogate.repr.prop.DF$geographic.held*100,na.rm=TRUE),2)` SD of the variation in geographic position among the planning units occupied by the species. On the other hand, to secure at least `r raptr.params.LST[[MODE]]$scenario.analysis$genetic.target*100` % of each species' adaptive genetic variation, prioritisations only needed to sample `r round(mean(surrogate.repr.prop.DF$environmental.held*100,na.rm=TRUE),2)` % $\pm$ `r round(sd(surrogate.repr.prop.DF$environmental.held*100,na.rm=TRUE),2)` SD of the environmental variation in planning units occupied by the species. Although to capture evolutionary processes, real-world conservation planning exercises may need higher surrogate targets.

<!-- adaptive genetic variation secured in prioritisations-->
Our results suggest that conservation planners can secure a reasonably representative sample of the intra-specific adaptive genetic variation using only amount-based targets. For most species, the spatial patterns of adaptive genetic variation clustered into a few distinct groups (eg. _Carex semipervens_ and _Cerastium uniflorum_; Figures S33 and S34 respectively). As a consequence, a prioritisation would only need to select a single planning unit in each cluster to secure a representative sample. This mechanism explains why adaptive variation accumulated much quicker than neutral genetic variation in the randomly generated prioritisations. Additionally, this mechanism also explains why the single-species prioritisations generated using only amount targets were often able to achieve high representation--they were essentially a random selection and so frequently contained planning units in different parts of the species' ranges (Figure 4a). Overall--and contrary to our original hypotheses--it seems that conventional reserve selection methods may secure a representative sample of species' adaptive genetic variation.

<!--  limitations/ weaknesses -->
There are several limitations associated with our analysis. Firstly, the size of the planning units we used (23 $\times$ 25 km) is much larger than typically used in regional conservation planning scenarios (1--10 km$^2$). We used this resolution because the genetic data were collected at this scale. Whilst we could have interpolated the genetic data to a finer resolution and used smaller planning units, this would have biased our analysis by artificially introducing additional spatial auto-correlation. Secondly, we used geographic distances as surrogates for neutral genetic variation, but connectivity in this landscape may be better described using distances that incorporate information on topography. However, such distances require species-specific scaling to adequate describe their connectivity (eg. [@r497]), and so cannot easily be utilised in large-scale planning exercises. Thirdly, we used amplified fragment length polymorphism data (AFLP; [@r453]) to characterise the main gradients of genetic variation. Whilst next-generation sequencing provides higher resolution genetic information than AFLP data ([@r480]), no multi-species genomic data set with similar properties currently exists, and our methodology would still have utilised only the main gradients of the genetic variation to capture broad-scale evolutionary processes.

<!--  conclusion -->
Broad-scale environmental and geographic variables are generally effective surrogates for representing adaptive and neutral genetic variation for most species in our study system. Moreover, data to calculate the surrogate variables are freely available for any location in the world. Careful use of such surrogates in conservation planning could vastly improve the chances of long-term biodiversity persistence for relatively little additional cost.
